---
  title: "Interactive web applications with RShiny"
author: "Kelly Hondula"
output: 
  md_document:
  variant: markdown_github
---
  
  # Interactive web applications in R
  
  This lesson presents an introduction to creating interactive web applications using the Shiny package in R. We will cover the following concepts:
  
  - [file structure template](#app-template)
  - [input objects](#input-objects)
  - [output objects](#output-objects)
  - [create a reactive object](#create-reactive-object)
  - [customizing layout design](#design-layout)
  - [upload or download files](#download-button)
  - [deploying your app](#how-to-deploy)
  
Make sure the shiny package is installed and loaded

```{r message=FALSE, warning=FALSE}
# library(shiny)
```

## App template

A shiny app consists of two components:

* **ui** which defines the user interface with the main function `fluidPage()`
* **server** which defines the instructions for assembling components of the app as a function of input and output objects

There are two ways to structure an app:

1. Create one file called `app.R`. In this file, define objects `ui` and `server` with the assignment operator `<-` and then pass them to the function `shinyApp()`. 
2. Split the template into two files named `ui.R` and `server.R`. 

When the `shiny` package is installed and loaded, RStudio will identify this file structure and create a green arrow with a "Run App" button when you open a file in the app. Applications use your current R session and are displayed in a browser window.

Test this by looking at an example within the shiny package:

```{r}
# runExample("01_hello")
```

Notice the stop sign that appears in the Console window while your app is running. This is because it is using the current R session. Make sure to end the app by either closing the external window or clicking the stop sign. 

## Input objects

The user interface and the server file interact with each other through __*Input()__ and __*Output()__ functions. 

Input objects collect information from the user and saves them based on the name given to the `inputId =` argument. Input objects also have a `label =` which is displayed to the user, as well as other values based on the type of input widget. 

Input objects are **reactive** which means that an update to this value by a user notifies the server file that its value has been changed. 

Input objects are stored in a list and are therefore referred to in the server file as `input$inputID`. 

Choices for inputs can be named using a list to match the display name to the value
# Male = "M", Female = "F", Both = "M" & "F"

Selectize input useful for long drop down lists

```{r}
# plots <- read.csv("../Data/plots.csv", stringsAsFactors = FALSE)
# species <- read.csv("../Data/species.csv", stringsAsFactors = FALSE)
# surveys <- read.csv("../Data/surveys.csv", na.strings = "", stringsAsFactors = FALSE)
# 
# # input object to select the taxa with radio buttons or check boxes
# # use input to filter data and create a plot
# # plot how many total of that taxa in each year
# 
# bird_species <- filter(species, taxa == "Bird")$species_id #Bird is input$pick_taxa
# surveys_birds <- filter(surveys, species_id %in% bird_species)
# barplot(table(surveys_birds$year))
```

Exercise: add an input feature to choose males, females, or both
hint: use either radio buttons or checkboxGroup input

## Output objects

Output objects are created through the combination of pairs of `render*()` and `*Output()` functions. Define a list of output objects in the server file using render functions:

```{r}
# output$my_plot <- renderPlot({})
```

Render functions build an object to display. The code inside the body of the render function will run whenever a reactive value inside the code changes.

Use the output ID to refer to that output element in the user interface file to place it in the app. 

```{r}
# ui.R
# fluidPage(
#   plotOutput("my_plot")
#   )
```


| render | output | description |
|--------|--------|-------------|
| renderPlot() | plotOutput() | plots |
| renderPrint() | verbatimTextOutput() | text |
| renderText() | textOutput() | text |
| renderTable() | tableOutput() | static table |
| renderDataTable() | dataTableOutput() | interactive table |
| renderUI() | uiOutput() | reactive input object |

Note that you can change the size of plot outputs by defining the number of pixels

## Reactive objects

Input objects that are used in multiple render functions to create different output objects can be created independently as **reactive** objects. This value is then cached to reduce computation required, since only the code to create this object is re-run when input values are updated. Use the function `reactive()` to create reactive objects and use them with function syntax, i.e. with `()`. 

make the filtered data set a reactive object to use in the plot and in another place

## Design layout

Page
Layout or rows
Panels, objects, or columns
Elements

The user interface file or object is where the page layout is defined. One option for the main function is `fluidPage()` and it takes arguments that are layout functions. Organize elements within defined layouts such as `fluidRow()`, `flowLayout()`, `sidebarLayout()`, `splitLayout()`, or `verticalLayout()`. 

Elements can be layered on top of each other using `tabsetPanel()`, `navlistPanel()`, or `navbarPage()`. 

Input and output objects in the user interface file must be separated by commas. 

change layout so there are 2 plots displayed together
add a tab with the data in a data table


### Customize appearance

Add images by saving those files in a folder called **www**. Link to it with img(src="<file name>")

Use html wrapper functions to add headers and formatted text

Add text that is a title for the graph with the appropriate selected taxa in it
Exercise: Below the graph, add text that says what year had the maximum number of that taxa recorded

Include markdown files

## Upload or download files

add a button to download a csv of the data or an image of the graph

## Deploy your app


## Shiny extensions

There are many ways to extend the functionality and sophistication of Shiny apps using existing tools and platforms. 

Leaflet

Plot.ly

## Additional references
